\newpage


# Appendix

We report some additional descriptive and robustness check. For additional statistics and for the entire R Code used generate this manuscript. Please visit this Github page []

## In- and Out-Degree distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE,out.width="50%"}


hist(friendship_out_w1,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=7,
     main="Friendship Out-degree distribution in wave 1",
     xlab="Out-degree",
     col="black")

hist(friendship_in_w1,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=16,
     main="Friendship In-degree distribution in wave 1",
     xlab="In-degree",
     col="black")

hist(friendship_out_w2,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=7,
     main="Friendship Out-degree distribution in wave 2",
     xlab="Out-degree",
     col="black")

hist(friendship_in_w2,
    xlim=c(0,15),
    ylim=c(0,50),
    breaks=16,
    main="Friendship In-degree distribution in wave 2",
    xlab="In-degree",
    col="black")


hist(t_out_w1,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=7,
     main="Trust Out-degree distribution in wave 1",
     xlab="Out-degree",
     col="black")

hist(t_in_w1,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=16,
     main="Trust In-degree distribution in wave 1",
     xlab="In-degree",
     col="black")

hist(t_out_w2,
     xlim=c(0,15),
     ylim=c(0,50),
     breaks=7,
     main="Trust Out-degree distribution in wave 2",
     xlab="Out-degree",
     col="black")

hist(t_in_w2,
    xlim=c(0,15),
    ylim=c(0,50),
    breaks=16,
    main="Trust In-degree distribution in wave 2",
    xlab="In-degree",
    col="black")
```

\newpage

## ERGMs 

### Model comparison 

<!-- \blandscape -->

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

stargazer::stargazer(ergm_1_w1,ergm_1_w2,
  ergm_2_w1,ergm_2_w2,
  ergm_3_w1,ergm_3_w2,
  ergm_4_w1,ergm_4_w2,
  ergm_5_w1,ergm_5_w2,
  dep.var.labels=c("W 1 ","W 2",
  "W 1 ","W 2",
  "W 1 ","W 2",
  "W 1 ","W 2",
  "W 1 ","W 2"
),
  font.size="small", 
  column.sep.width = "1pt",
  title = "All fitted ERGMs models for friendship (waves comparison)",
  header=FALSE,
  type="latex",
  float.env = "sidewaystable")

```

<!-- \elandscape -->

\newpage 

### Model Diagnostics

We only report the diagnostics for the selected model (Model 4)


```{r,echo=FALSE, message=FALSE, warning=FALSE, out.width="50%", fig.cap="MCMC diagnostics Wave 1", fig.pos="H",results='hide', fig.align="center"}

mcmc.diagnostics(ergm_4_w1)

```


```{r,echo=FALSE, message=FALSE, warning=FALSE,out.width="50%", fig.cap="GOF Wave 1", fig.pos="H",fig.align="center"}

plot(gof(ergm_4_w1))

```


```{r,echo=FALSE, message=FALSE, warning=FALSE,out.width="50%",fig.cap="GOF Wave 2",fig.pos="H",fig.align="center"}

plot(gof(ergm_4_w2))

```

\newpage

## SIENA 

### Uniplex GOF

```{r,cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE,fig.pos="H",results='hide',fig.align="center"}

plot(gof.fr.indegree)
plot(gof.fr.outdegree)
plot(gof.fr.geodesic)
plot(gof.fr.triadcensus, center = TRUE, scale = TRUE)

```


